---
title: "Stat c131a Final Project"
author: Mia Patzakis
format: html
---


```{r}
#| message: FALSE
install.packages("broom")
install.packages("GGally")
library(dplyr)
library(tidyverse)
library(broom)
library(GGally)
```


# Background

Primary biliary cholingitis (PBC) is a chronic disease where someone's bile ducts in the liver become inflamed and damaged. The bile ducts transport bile from the liver to the small intestine, where it helps digest food. When the bile ducts become inflamed and lose function, bile can build up in the liver and cause liver damage and even liver failure, causing symptoms like abdominal pain, jaundice, and swelling. Because this condition involves tissue being destroyed by one's own body, it is considered an autoimmune disease. PBC is not incredibly common, with around 200,000 people affected each year in the United States. It is more common in women than men - only 10% of cases can be attributed to men. Some risk factors for PBC include being female, being between the ages of 30 and 60, being genetically predisposed to getting the condition, getting frequent urinary tract infections and possibly some behaviors like smoking.

Survival analysis is a type of "time-to-event" analysis, where the variable "time-to-event" is the time until an event of interest occurs. In the survival context, the event of interest is an adverse event such as death or a heart attack. Thus survival analysis analyzes the amount of time until death occurs. Survival analysis is applicable for many medical studies, such as examining how long a patient can live with a certain disease or condition, comparing survival rates in different clinical trial groups, and other questions regarding rates of survival due to different factors or past a given time. It is thus particularly prominent for studying cancer prognoses. For example, one may study the survival time from relapse or remission, or survival time after starting a particular treatment.

This project may be helpful for the "real world" because the number of days a person survives with PBC is influenced by many factors, like age, different biomarkers, or drugs, and there is inconclusive evidence on how these factors affect PBC mortality. This project demonstrates how researchers could use predictive methods like multivariate linear regression to pinpoint relationships between these explanatory variables and number of days someone can survive. Additionally, this project explores many different methods to predict an outcome, like random forests, decision trees and cross-validation, which can be helpful for research on survival analyses for not only PBC but other diseases.

# First Steps

```{r}
#| message: FALSE
cholangitis <- read.csv("cholangitis.csv")
# reading data into R

# creating levels for all categorical variables
status_levels <- c("C", "CL", "D")
drug_levels <- c("D-penicillamine", "Placebo")
sex_levels <- c("F","M")
ascites_levels <- c("Y", "N")
hepatomegaly_levels <- c("Y", "N")
spiders_levels <- c("Y", "N")
edema_levels <- c("N", "S", "Y")
stage_levels <- c("1","2","3","4")

# adding factored categorical variables in place of nonfactored variables
cholangitis$status <- factor(cholangitis$status, levels=status_levels)
cholangitis$drug <- as.character(factor(cholangitis$drug, levels=drug_levels))
cholangitis$sex <- factor(cholangitis$sex, levels=sex_levels)
cholangitis$ascites <- factor(cholangitis$ascites, levels=ascites_levels)
cholangitis$hepatomegaly <- factor(cholangitis$hepatomegaly, levels=hepatomegaly_levels)
cholangitis$spiders <- factor(cholangitis$spiders, levels=spiders_levels)
cholangitis$edema <- factor(cholangitis$edema, levels=edema_levels)
cholangitis$stage <- factor(cholangitis$stage, levels=stage_levels)
cholangitis$cholesterol <- as.integer(cholangitis$cholesterol)


cholangitis <- cholangitis %>% mutate(drug = replace(drug, is.na(drug), "NA"))
# creating new category for NA drugs

cholangitis <- cholangitis %>%
  filter(!is.na(tryglicerides) & !is.na(cholesterol))
# getting rid of NA values in categorical columns (only removes 5 rows)
```

In the above code I've loaded our cholangitis data into R studio and turned the categorical variables into factors. I also turned the "NA" values in the drug vector into it's own category of drug and removed numerical NA values. 


```{r, out.width="60%", out.height="60%"}
#| message: FALSE
# creating a pairs plot for first first set of variables with n_days
ggpairs(cholangitis[,c("n_days", "drug", "ascites", "bilirubin", "albumin", 
                       "copper", "alk_phos")], aes(color = drug)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text.y = element_text(size = 7, angle = 0)) +
  ggtitle("Drug NA category associated with lower n_days")


```


```{r, out.width="60%", out.height="60%"}
#| message: FALSE
# pairs plot for second set of variables with n_days
ggpairs(cholangitis[,c("n_days", "sex", "age", "spiders", "cholesterol",
                       "sgot", "tryglicerides")], aes(color=sex),
        ) + theme(axis.text.x = element_text(angle = 45, hjust = 1),
                  axis.text.y = element_text(size = 7, angle = 0)) +
  ggtitle("Age is more negatively associated with n_days for men than women")


```

```{r, out.width="60%", out.height="60%"}
#| message: FALSE
# pairs plot for last set of variables with n_days
ggpairs(cholangitis[,c("n_days", "hepatomegaly", "edema", "platelets",
                       "prothrombin", "stage")], aes(color=hepatomegaly)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text.y = element_text(size = 7, angle = 0)) +
  ggtitle("Hepatomegaly positively associated with less n_days than no hep.")

```

For the first step in my EDA I made a pairs plot to visualize the categorical and numerical variables against one another and how they correlate with n_days. Because there are so many explanatory variables, I created three pairs plots all with a different combination of categorical and numerical variables in three different code chunks so that the plot doesn't get too cramped. Below is more EDA and my analysis of these plots.


```{r, out.width="60%", out.height="60%"}
# selecting all continuous variables to create correlation heat map
numeric_data <- cholangitis %>%
  select_if(is.numeric)

cor_matrix <- as.data.frame(cor(numeric_data))

cor_matrix %>%
  rownames_to_column("Variables_1") %>%
  pivot_longer(-c(Variables_1), 
               names_to = "Variables_2", 
               values_to = "Correlation") %>%
  arrange(Correlation) %>%
  mutate(Variables_1 = factor(Variables_1, levels = unique(Variables_1)),
         Variables_2 = factor(Variables_2, levels = unique(Variables_2))) %>%

  ggplot(mapping = aes(x = Variables_1, y = Variables_2)) + 
    geom_tile(aes(fill = Correlation)) +
    scale_fill_gradient(low = "black", high = "red",
                        breaks = seq(-1, 1, by = 0.2)) +
    scale_x_discrete(guide = guide_axis(n.dodge = 1)) +
  scale_y_discrete(guide=guide_axis(n.dodge=1.5)) +
    theme_classic() +
    theme(axis.title.x = element_blank(),
          axis.title.y = element_blank(),
          axis.text = element_text(size=6),
          axis.text.x=element_text(angle=45,hjust=1)) +
  ggtitle("Most variables weakly or not correlated with one another")
```

Here I've selected all of the numerical continuous variables to create a correlation heat map to see to what degree each variable correlates with each other. I've also included the "ID" variable so that we can see if the ID numbers also correlate with the other variables in any notable way. We can see from the heat map that the ID variable does not correlate significantly with the other variables.

After performing EDA, we can see that all of the numerical variables are relatively symmetrical and normally-distributed, as shown from the density plots in each pairs plot. The variables bilirubin, copper, cholesterol and prothrombin, seem to be skewed-right. Many variables, as apparent from the pairs plot scatter plots, are clustered close together with a few outstanding points. This suggests that we may need to log transform these variables. From our correlation matrix heat map (and our pairs plots) we can see that most of the explanatory variables have a weakly negative or no correlation with n_days survived. Most of the explanatory variables are also not correlated with one another. Bilirubin seems to weakly correlate with most continuous variables like sgot, cholesterol, promthrombin and tryglicerides. Albumin seems to have the highest correlation with n_days. As we can see from the second pairs plot, there are many more females than males in the dataset, which is consistent with the fact that PBC is more common in women than men. 

# Exploratory Modeling

```{r, out.width="50%", out.height="50%"}
#| message: FALSE
# calculating cooks distances using the filtered data frame
cholangitis$cooks <- cooks.distance(lm(n_days ~ ., 
                                       data = cholangitis))

ggplot(cholangitis, aes(x=id,
                     y=cooks)) +
  geom_line() + xlab("Observation Number") +
  ylab("Cook's Distance") + 
  ggtitle("Few points over 0.03 Cook's Distance")

# filtering out ids above 0.03 cooks distance
cholangitis <- cholangitis %>% filter(cooks < 0.03)
```

In the above code chunk, I've created a Cook's distance plot to see how much the coefficients in the linear model will change if each observation is left out, so as to see if I should exclude any points from the regression model that have too large an influence on the prediction. We can see that the Cook's distance is very large for a few points, so I decided to exclude points with a distance of over 0.03 and reassign the cholangitis variable without them. After removing these points, I moved on to see if I should transform any variables.

```{r, out.width="50%", out.height="50%"}
#| message: FALSE
#| echo: FALSE
# checking to see if I should transform bilirubin
ds1 <- ggplot(cholangitis, 
       aes(x=bilirubin, y=n_days)) + 
  geom_point() + ggtitle("Bilirubin vs. n_days is unevenly clustered")


ds2 <- ggplot(cholangitis, 
       aes(x=log(bilirubin), y=n_days)) + 
  geom_point() + ggtitle("Bilirubin logged has slight negative correlation with n_days")

```

```{r, out.width="50%", out.height="50%"}
#| message: FALSE
#| echo: FALSE
# checking to see if I should transform alk_phos
ds3 <- ggplot(cholangitis, 
       aes(x=alk_phos,
           y=n_days)) + 
  geom_point() + ggtitle("Alk_phos vs. n_days is unevenly clustered")


ds4 <- ggplot(cholangitis, 
       aes(x=log(alk_phos),
           y=n_days)) + 
  geom_point() + ggtitle("Alk_phos logged is more homeoskedastic")

```

```{r, out.width="50%", out.height="50%"}
#| message: FALSE
#| echo: FALSE
# checking to see if I should transform cholesterol
ds5<- ggplot(cholangitis, aes(x=cholesterol, y=n_days)) +
  geom_point() + ggtitle("Cholesterol vs. n_days is unevenly clustered")

ds6<-ggplot(cholangitis, aes(x=log(cholesterol),y=n_days)) +
  geom_point() + ggtitle("Cholesterol logged vs. n_days is less skewed")

```

```{r, out.width="50%", out.height="50%"}
#| message: FALSE
#| echo: FALSE
ds7<-ggplot(cholangitis, aes(x=copper, y=n_days)) +
   geom_point() + ggtitle("Copper vs. n_days is unevenly clustered")


ds8<-ggplot(cholangitis, aes(x=log(copper), y=n_days)) +
   geom_point() + ggtitle("Copper logged has slight negative correlation with n_days")

```

I decided to log transform the variables bilirubin, cholesterol, copper, alk_phos, sgot, tryglicerides, and prothrombin because these variables are all skewed right and non-homoskedastic before being log-transformed. I will keep these variables logged for the rest of my modeling and prediction because these transformations clarify the relationships between these variables and n_days. 

```{r}
# putting together the model with all factors except id
first_model <- lm(n_days~ status + drug + age+sex+ascites+ 
                  hepatomegaly+ spiders+ edema+ log(bilirubin)+
                  log(cholesterol)+ albumin+ log(copper)+ log(alk_phos)+
                  log(sgot)+log(tryglicerides)+platelets+ log(prothrombin)+
                  stage, data = cholangitis)

summary(first_model)
```

Above is the model I created to predict n_days with all of the factors in the data set, before I decide how to subset my variables. As aforementioned, I excluded observations with a cook's distance above 0.03, and kept certain variables log-transformed. The summary of this model shows that variables like bilirubin logged and copper logged and others with low p-values are nonzero given the inclusion of the other variables. 


```{r, out.width="50%", out.height="50%"}
#| message: FALSE
# Diagnostics for this model

# creating residuals and predicted values
cholangitis$resids <- resid(first_model)
cholangitis$first_predictions <- fitted(first_model)
# creating residuals variable with all-variables model

cholangitis %>%
  ggplot(mapping = aes(x = first_predictions, y = resids)) +
  geom_point(alpha = .5, color = "red") +
  geom_line(y=0) +
  theme_classic() + xlab("Predictions") +
  ggtitle("Residuals from first MLR model show no pattern")

# creating qqplot 
cholangitis %>%
  ggplot(mapping =aes(sample=resids)) + 
  stat_qq() + stat_qq_line() + 
  theme_minimal() +
  ggtitle("Residuals deviate slightly from qqplot")
```

To run diagnostics for this initial model that includes all of the explanatory variables, I created a residuals vs. predicted values scatter plot to see if the residuals violate any of the assumptions that they should have a mean of 0, no trend or shape, and a constant variance. As we can see from the above plot, the mean of the residuals does appear to be 0, and they do not appear to follow a trend, so these regression assumptions are met. The variance does not appear to be constant, however, as the center clump of residuals have a smaller variance compared to residuals above 0 and on the edges of the cluster, so the residuals do not appear to meet that assumption. I also created a qqplot to check the assumption of residuals being normally distributed. The residuals follow the qqplot fairly well, but deviate upward at the right tail, suggesting a slight departure from normality.


```{r, out.width="50%", out.height="50%"}
# what drugs are associated with most number of days?
ggplot(data =cholangitis, 
       mapping = aes(x = n_days, 
                     fill = drug)) +
  geom_boxplot() + ggtitle("Drug NA cateogry associated with lowest n_days")
```

According to our pairs plot from earlier (from which I've included a larger box plot above), there visually appears to be no significant difference between D-penicillamine or the placebo on the number of days survived. However, according to the drug placebo coefficient in the linear model, someone receiving the placebo is expected to have n_days 22.8 lower than with D-penicillamine, assuming all other variables are held constant. However, this coefficient has a high p-value, so there isn't strong evidence to suggest a significant difference in n_days between individuals with the placebo and with D-penicillamine.

However, the "NA" drug category visually seems to span a lower range of n_days than the placebo and D-penicillamine. In the linear model, the estimated coefficient is -417, implying that individuals with DrugNA have n_days lower by 417 compared to individuals with D-penicillamine, holding other variables constant. The p-value for this estimate is very small, suggesting there is a significant notable impact on this drug category and lowering the outcome of n_days. Because this category is "NA", and there is no difference in n_day outcome between the drug and the placibo, there is little evidence from this data that the drug has an impact on number of days survived in this study. 

# Predictive Modeling

```{r}
# making new df with log transformed vars
# "cholangitis-transformed"
ct <- cholangitis %>% mutate(
  bilirubinL = log(bilirubin),
  cholesterolL = log(cholesterol),
  copperL = log(copper),
  alk_phosL = log(alk_phos),
  sgotL = log(sgot), 
  trygliceridesL = log(tryglicerides),
  promthrombinL = log(prothrombin)
) %>% #taking out the old non transformed and unneeded variables
  select(-c("bilirubin", "cholesterol", "copper",
            "alk_phos", "sgot", "tryglicerides", 
            "prothrombin", "cooks",
            "id", "resids", "first_predictions") 
  )
```

Here, I've made a new data frame "ct" (cholangitis transformed) with the log transformed variables replacing the old variables. I have also gotten rid of the columns, "cooks", "id", "resids", and "first_predictions" so that they do not interfere with my model as they are not explanatory variables.

```{r}
# train test split
set.seed(131)
train_pct <- .7
indices <- seq(from = 1, to = nrow(ct), by = 1)
training_indices <- sample(x = indices, replace = FALSE, 
                           size = nrow(ct)*train_pct)
chol_train <- ct %>%
  slice(training_indices)

chol_test <- ct %>%
  slice(-training_indices)
```

The above code contains my train-test split. I decided to use a lower ratio split of 0.7 so that the training set isn't too large and we have a substantial amount of data to test on. Too large of a training set can train the model to fit too well (over-fit) to the training set's nuances and not model the patterns of the testing set accurately, especially because there are so many variables being included in this model. 

```{r}
#| message: FALSE
# installing and loading the leaps package 
install.packages("leaps")
library(leaps)
```

```{r}
# variable selection
best_k_model_results <- regsubsets(n_days ~ ., data = chol_train, nvmax = 23)

summary(best_k_model_results)
summary(best_k_model_results)$rsq
```

```{r, out.width="50%", out.height="50%"}
# plotting the number of variables included against rsq values
rsq_df <- data.frame(num_vals = seq(1, 23, by = 1), 
                 rsqs = summary(best_k_model_results)$rsq)

ggplot(rsq_df, aes(x=num_vals, y=rsqs)) + geom_point() +
  scale_x_continuous(breaks = seq(1, 23, by = 1)) +
  theme_minimal() + xlab("Number of Variables Included") +
  ylab("RSQ values") +
  ggtitle("RSQs level out with inclusion of 11 variables")

```

I used a stepwise variable selection method to see which variables have the biggest impact on best predicting n_days as more variables are included in the model. As we can see from the rsq output and the graph above that shows the impact of the number of variables on the rsq for predicting n_days, adding more variables than 11 starts to have little impact on improving the rsq. We can now compare the root mean squared error of models with different amounts of variables to confirm this information and what variables to include in the final model. 

I selected different amounts of variables based on decreasing importance to create different models to test and see which one has the lowest RMSE. From this stepwise variable selection, I derived five different models to test:


```{r}
# creating prediction vectors for each of the linear models using test data
formulas <- list(
  n_days ~ bilirubinL + albumin,
  n_days ~ bilirubinL + albumin + alk_phosL + status,
  n_days ~ bilirubinL + albumin + alk_phosL + status + drug + stage + edema + promthrombinL,
  n_days ~ bilirubinL + albumin + alk_phosL + status + drug + stage + edema + promthrombinL + copperL,
  n_days ~ bilirubinL + albumin + alk_phosL + status + drug + stage + edema + promthrombinL + copperL + trygliceridesL)
# fitting each model
models <- lapply(formulas, function(f) lm(f, data = chol_train))
predictions <- lapply(models, predict, newdata = chol_test)
#RMSE's for each model
observations <- chol_test %>% select(n_days) %>% pull()
RMSEs <- sapply(predictions, function(pred) sqrt(mean((pred - observations)^2)))
cat(paste0("RMSE for Models: ", paste(RMSEs, collapse = " | "), "\n"))

```

Above I found the RMSE for each model I created by using our test data to predict new values of n_days and finding the error with the actual observed n_days observations. Because the RMSE is smallest for the fourth model, that is the subset of variables that I will include, because the error is minimized when predicting the outcome n_days on the test set. 

My linear model thus includes the variables bilirubin (logged), albumin, alk_phos (logged), status, drug, stage, edema, promthrombin (logged), and copper (logged) to predict n_days.


```{r}
# loading rpart and plot libraries
library(rpart)
library(rpart.plot)
```

```{r}
# constructing the decision tree model using all variables 
decision_tree <- 
  rpart(formula = n_days ~ ., data = chol_train)

# displaying the tree
rpart.plot(decision_tree, type = 3)

decision_tree

printcp(decision_tree)

```

Above I fit a regression tree with all of the explanatory variables. In this first pass at a regression tree, the variables being used to make splits are bilirubin (logged), drug, stage, albumin, alk_phos (logged), promthrombin (logged), age, status, and sgot (logged). 

From the printed cp output, we can see that the lowest xerror value while keeping a low cp, maintaining the lowest split number possible, comes with five splits. The xerror is 0.78145 and the cp value for a five-split tree is 0.028438. Because it is worth having more tree branches for a smaller cp value, we will use the tree with split number five, because it minimizes the number of splits while maintaining a small cp and xerror. Thus we can create our final decision tree model with the new specified cp value:

```{r}
# new decision tree model with lower cp
dt_model <- rpart(formula = n_days ~ ., data = chol_train, 
                  cp = 0.028438)
```

```{r}
rpart.plot(dt_model, type = 3)
```
In this new "pruned" tree model with a cp of 0.028, the variables bilirubin (logged), albumin, alk_phos (logged), stage, and promthrombin (logged) are being used to make splits. Now we can move on to make a random forest model. 


```{r}
#| message: FALSE
# loading random forest package
library(randomForest)
```

```{r}
# creating our random forest model with all variables
random_forest <- 
  randomForest(formula = n_days~., 
        data = chol_train, # creating model with train data
        cutoff = c(0.5, 0.5),
        importance = TRUE)

importance(random_forest)
```


```{r}
varImpPlot(random_forest, 
           main = "Variable Importance Plot",
           pch = 16)
```

Above I fit a random forest model using all explanatory variables to predict n_days. The default number of trees fit in the random forest function is 500. From our random forest importance output and variable importance chart, we can see that bilirubin (logged) and albumin are considered the most important variables because these have some of the highest values both %IncMSE and IncNodePurity.

```{r}
# random forest prediction object 
rf_predictions <- predict(object = random_forest, newdata = chol_test)

# decision tree prediction vector
dt_predictions <- predict(object = dt_model, newdata = chol_test)
```

```{r}
# calculating RMSE for each model

MSE_dt <- mean((dt_predictions-observations)^2)
# mse for decision tree model
sqrt(MSE_dt)

MSE_rf <- mean((rf_predictions-observations)^2)
# mse for random forest model
sqrt(MSE_rf)
```

In the above three code chunks, I output prediction vectors for both my decision tree model and my random forest model so that I could calculate a RMSE for each model, using the observed n_days vector. We can see from the RMSE values that the random forest model is more accurate than the decision tree model.



| Model              | RMSE       |
|-------------------- |------------|
| Linear Regression   | 850.9564      |
| Regression Tree    | 1028.239       |
| Random Forest       | 894.4369       |


When we compare the RMSE's across all of these models, we can see that it is lowest for the MLR model I chose. This suggests that this model performed the best on the test data, meaning that it captured the trends of the cholangitis training data without overfitting in comparison to the decision tree and random forest models I constructed. This is thus the model I will use to predict the number of days of survival after registration for the study.
